Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-12 vertical-thermal-review.Rmd
form 2 new VOC paragraphs

Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10˚C) during sunflecks [@waySunflecksTreesForests2012].
In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].

<!--VOCs-->
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017]. 
Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].

Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein]. 
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.

>>>>

Biochemical protection against foliage light and heat damage increase with peak radiation loads, and thus tend to be higher in the upper canopy than in the understory (Table `r table_leaf_traits`). 
Further, by reducing photosynthesis, heat and associated stomatal closure from VPD can indirectly reduce the capacity to use light energy for photochemistry, thereby increasing the potential for photoinhibition.
Accumulation of excess light energy results in overreduction of the electron transport chain and the formation of harmful reactive oxygen species [@niyogiSafetyValvesPhotosynthesis2000; @suzukiReactiveOxygenSpecies2006a].
A ubiquitous defense is a rapidly inducible non-photochemical quenching (NPQ) mechanism that responds to the increased thylakoid pH gradient caused by accumulated light energy [@niyogiSafetyValvesPhotosynthesis2000; @gossBiodiversityNPQ2015].
This form of NPQ is accomplished by the interconversion of xanthophyll cycle pigments—violaxanthin, antheraxanthin, and zeaxanthin (VAZ)—which regulates the capacity for de-excitation of chlorophyll through thermal dissipation instead of photochemistry.
Due to higher leaf temperatures and hydraulic limitations, upper canopy leaves spend more time with stomata closed, and therefore require a high capacity to dissipate excess light energy [@niinemetsPhotosynthesisResourceDistribution2007].
Correspondingly, leaves in higher light environments show a greater capacity for NPQ and higher concentrations of VAZ as well as other carotenoids (e.g., beta carotene and lutein) employed as antioxidant defenses [Table `r table_leaf_traits`, @niinemetsAcclimationHighIrradiance1998; @garcia-plazaolaAcclimationAntioxidantPools2004a; @scartazzaInvestigatingEuropeanBeech2016a].
In contrast, leaves deeper in the canopy experience rapid light and temperature surges (can exceed 10$^\circ$C) during sunflecks [@waySunflecksTreesForests2012].
In order to maximize carbon uptake and minimize damage during sunflecks, shade-acclimated leaves tend to induce photochemical processes more quickly [@urbanInductionPhotosynthesisImportance2007], and also show a steeper response of NPQ to light compared to sun-acclimated, upper canopy leaves [@scartazzaInvestigatingEuropeanBeech2016a].

<!--VOCs-->
Capacity for heat-sensitive, light-dependent (photosynthetically linked) emissions of volatile organic compounds (VOCs), including isoprene and monoterpenes, enhances photosynthetic thermal tolerance by regulating antioxidant defenses and other metabolic processes [@sharkeyIsopreneEmissionPlants2008; @taylorCapacityEmitIsoprene2019; @monsonLeafIsopreneEmission2021; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005; @riedlmeierMonoterpenesSupportSystemic2017]. 
Emission of the VOCs isoprene and a diversity of monoterpenes are similarly light and temperature sensitive, and play similar roles in photosynthetic thermal tolerance [@taylorCapacityEmitIsoprene2019; @sharkeyIsopreneEmissionPlants2008; @vickersUnifiedMechanismAction2009; @copoloviciCapacityThermalProtection2005]. 
In contrast to VOCs stored in oils that are released passively by heat and wounding, such as the monoterpenes responsible for pine scent, light-dependent emissions are linked to photosynthetic substrate supply and are dynamically tuned to environmental and metabolic conditions as a component of metabolic regulatory processes [@monsonLeafIsopreneEmission2021; @laothawornkitkulBiogenicVolatileOrganic2009; @lantzIsopreneNewInsights2019; @riedlmeierMonoterpenesSupportSystemic2017]. 
Due to its much higher emission rates and relative ease of detectability, isoprene has received far more study in the field, though both isoprene and monoterpenes are expressed by a wide diversity of angiosperms and gymnosperms across all biomes of the globe [see @taylorNewFieldInstrument2021 and references therein].

Less than half of tree species express significant light-dependent emissions [@kesselmeierBiogenicVolatileOrganic; @taylorIsopreneEmissionStructures2018], such that variation in emissions across the vertical profile can be influenced both by vertical variation in species compositions and plasticity in emission rate capacities [@taylorNewFieldInstrument2021].
Within species, isoprene emission rates tend to increase toward brighter and hotter microenvironments [@niinemetsPhotosynthesisResourceDistribution2007], and across landscapes, emitting species increase in relative abundance toward hotter climates, exceeding half of trees in warm tropical forests [@taylorIsopreneEmissionStructures2018]. 
However, a recent study found a contrasting interspecific vertical structuring of emission capacities, with more emitting species and higher species-maximum emission rates in the mid-canopy region of an Amazonian forest [Table `r table_leaf_metabolism`, @taylorNewFieldInstrument2021]. 
Similarly, within tree crowns of European beech, monoterpene emissions were found to be highest in semi-shaded leaves beneath the canopy surface [Table `r table_leaf_metabolism`, @simpragaVerticalCanopyGradient2013]. 
This pattern may indicate the importance of temporal variability in thermal conditions as distinct from the long-term average. 
Temperature sensitive VOC emissions have been hypothesized to enable real-time acclimation to rapidly changing leaf thermal environments typical of the mid-canopy region [see section 1, @sharkeyIsopreneEmissionPlants2008]. 
Future work seeking to understand how temporal dynamics of leaf heating [see @leighInfluenceLeafSize2017] with height and influence leaf function will be important for resolving the role of VOCs in forest thermal sensitivity across the vertical gradient. 
Given current understanding, we may hypothesize that light-dependent VOC emissions are important for mid-canopy thermal responses, while other traits play the same role in sun-exposed canopy leaves.
